Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-12-22 threshold/draft.Rnw
Minor changes to conclusion

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation. Lemma
\ref{sumlemma} decomposes Euclidean distance between two vectors in terms of
singular basis vectors and Algorithm \ref{prune} can be applied directly to the
problem of finding all pairs of vectors that meet a thresholded Euclidean
distance with small changes in steps 3, 4, and 7.  By equivalence of norms,
thresholded Euclidean distance can be extended to other distance metrics. The
reference \proglang{R} implementation \citep{sup} includes functions for thresholded
distance and correlation.

@@

We present a simple algorithm and reference \proglang{R} implementation based on a
truncated singular value decomposition (SVD) that efficiently computes
thresholded subsets of large correlation and distance matrices.  We show that
the natural connection between data covariance and the SVD can result in
substantial computational and memory savings, and illustrate these savings with
a few genomics data experiments.  Moreover, the main computational work in the
algorithm is easily parallelized and deployed across computational
clusters.

>>>>

The idea of evaluating threshold conditions by projection into low dimensional
subspaces applies to distance metrics similarly to correlation, see
\citep{svd-similarity} for a related approach. Lemma \ref{sumlemma} decomposes
Euclidean distance between two vectors in terms of singular basis vectors and
Algorithm \ref{prune} can be applied directly to the problem of finding all
pairs of vectors that meet a thresholded Euclidean distance with small changes
in steps 3, 4, and 7.  By equivalence of norms, thresholded Euclidean distance
can be extended to other distance metrics. The reference \proglang{R}
implementation \citep{sup} includes functions for thresholded distance and
correlation.

@@

We present a simple algorithm and reference \proglang{R} implementation based
on a truncated singular value decomposition (SVD) that efficiently computes
thresholded subsets of large correlation  matrices.  We show that the natural
connection between data covariance and the SVD can result in substantial
computational and memory savings, and illustrate these savings with a few
genomics data experiments.  The main computational work in the algorithm is
easily parallelized and deployed across computational clusters.
